---
title: "Phyloseq Preprocessing"
author: "Marian Schmidt"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/02_PhyloseqPreProcessing/") # send any figure output to this folder 
```


# Before you start

## Set my seed 
```{r set-seed}
# Any number can be chosen 
set.seed(238428)
```


# Goals for this file 

Here in this file, we are going to process the data to have our final data for alpha and beta diversity analyses! 

We will need to: 

1. Load in ASV table, tax table, and metadata. 
2. Clean & combine the data into one phyloseq object. 
3. Clean up the data and remove ASVs/Samples we do not need.  
      - Remove Control Samples (and their ASVs) *be careful!*  
      - Remove Chloroplasts.  
      - Remove Mitochondria.  
      - Remove samples that do not have "enough" sequencing. 
5. Write out the final phyloseq object to work with later. :) 


# Before you start

## Set my seed 
```{r set-seed}
# Any number can be chosen 
set.seed(238428)
```


## Load Packages
```{r load-packages}
# Efficient loading of the packages 
pacman::p_load(dada2, tidyverse, patchwork, phyloseq, install = FALSE)
```


## Load Data 

We need to load in: 

1. ASV Table 
2. taxonomy table 
3. Metadata  

### ASV Table
```{r load-asv-table}
# First, let's load in our feature count table
# Load the ASV counts that was generated in the 01_DADA2 workflow 
load("data/01_DADA2/ASV_counts.RData")
# THis data has the ASV names 
dim(asv_tab)
asv_tab <- t(asv_tab)
dim(asv_tab)
head(asv_tab)[,1:5]

# Fix names 
sample_names <- colnames(asv_tab)
samples_fixed <- sapply(strsplit(basename(sample_names), "_"), `[`,1) 

# Now add it to the names
colnames(asv_tab) <- samples_fixed


# Put in a double check 
stopifnot(dim(asv_tab) == dim(noChimeras_ASV_table))
```


### Taxonomy Table
```{r load-tax-table}
taxonomy_tab <- read.table("data/01_DADA2/ASV_taxonomy.tsv",  sep = "\t", skip = 1) 
head(taxonomy_tab)

colnames(taxonomy_tab) <- c("ASV", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species") 
head(taxonomy_tab)

# Assign to tax_table()
tax_mat <- taxonomy_tab %>%
  tibble::column_to_rownames(., var = "ASV") %>%
  as.matrix()
```

### Metdata 
```{load-metadata}
# And load our tracking data
load("data/01_DADA2/track_read_counts.RData") # loads track_counts_df
head(track_counts_df)

# Fix it up
track_reads <- 
  track_counts_df %>%
  tibble::column_to_rownames(., var = "names") 

head(track_reads) 
```

## Handoff to phyloseq 
```{r phyloseq-handoff}
raw_physeq <- phyloseq(otu_table(asv_tab, taxa_are_rows = TRUE),
                         sample_data(track_reads),
                         tax_table(tax_mat))
raw_physeq

save(raw_physeq, file = paste0("data/raw_physeq.RData"))
```





