---
title: "Assigning ASVs with DADA2"
author: "Mar Schmidt"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/02_AssignASVs/") # send any figure output to this folder 
```


# Goals of this file 

1. Infer Errors on forward and reverse reads individually.
2. Identy ASVs on forward and reverse reads separately, using the error model.
3. Merge forward and reverse ASVs into "contiguous ASVs".  
4. Generate the ASV count table
5. Trim ASV Lengths
6. Remove Chimeras. 
7. Assign Taxonomy with the Silva Database.
8. Write out relevant files: `asv_table`, `asv_fastas`, `tax_table`, and `sample_data`.

#### Input that we need: 

1. Filtered `fastq.gz` files generated from `01_QualityTrimming.Rmd`.
2. Sample name vector 

#### Output that we need: 

1. ASV Count Table: `asv_table`
2. ASV fasta file: for building phylogenetic trees later.
3. Taxonomy Table  `tax_table`
4. Sample Information: `sample_data`  track the reads lots throughout DADA2 workflow. 



# Setting up the environment

## Set the seed 
```{r set-seed}
# Any number can be chosen 
set.seed(238428)
```

### Timing of Script

I'd like to know how long this file took to run, which we will record at the end of the script. 

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
start_time
```


## Load Libraries 
```{r load-libraries}
# Efficient package loading with pacman 
# Don't forget to install pacman and DT if you don't have it yet. :) 
pacman::p_load(tidyverse, BiocManager, devtools, dada2, 
               phyloseq, patchwork, DT, install = FALSE)
```


## Load in Filtered fastqs

```{r load-filtered-fastqs}
# Place filtered reads into filtered_fastqs_path
filtered_fastqs_path <- "data/01_DADA2/02_filtered_fastqs"

# Intuition Check 
filtered_fastqs_path

# create 2 vectors: filtered_forward_reads & filtered_reverse_reads
filtered_forward_reads <- 
  list.files(filtered_fastqs_path, pattern = "_R1_filtered.fastq.gz",
             full.names = TRUE)  

# Intuition Check 
length(filtered_forward_reads)

# reverse reads
filtered_reverse_reads <- 
    list.files(filtered_fastqs_path, pattern = "_R2_filtered.fastq.gz",
             full.names = TRUE)  

# Intuition Check 
head(filtered_reverse_reads)

```

## Assign Sample Names 

```{r sample-names}
# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(filtered_forward_reads), "_"), `[`,1) 

# Intuition Check 
head(sample_names)
```



# Error Modelling 

## How does it work? 

This is the step along the workflow where we try to estimate what is a *mistake* from technological sequencing error versus what is *true biological variation*. Therefore, this step is another critical step that we need to be mindful of! This is also what makes DADA2 unique!  

Specifically, we will infer error rates for all possible *transitions* within purines and pyrimidines (A<>G or C<>T) and *transversions* between all purine and pyrimidine combinations. The error model is learned by alternating estimation of the error rates and inference of sample composition until they converge. It starts with abundant sequences first and then goes to less abundant sequences. It will:  

1. Starts with the assumption that the error rates are the maximum (takes the most abundant sequence ("center") and assumes it's the only sequence not caused by errors).  
2. Compares the other sequences to the most abundant sequence. 
3. Uses at most 10^8^ nucleotides for the error estimation. (Though, sometimes we increase this parameter in the case of binned sequencing quality scores.)  
4. Uses parametric error estimation function of loess fit on the observed error rates. 

## Important Notes

**First, every individual sequencing run needs to have it's own and separate error model!** This is because every sequencing run is unique—even if the flow cell and DNA/samples are *exactly the same*. 

While you can combine multiple datasets together after running them through DADA2, if you'd like to limit bias and combine data sets from multiple sequencing runs, it is ideal to do the following:  

1. Use the exact same DNA extractions and library prep (*e.g.* same primer sets). 
2. Perform the exact same `filterAndTrim()`, with equivalent parameters between sequencing runs. This is because we need to have the exact same ASV length expected by the output of our filtering and trimming.
3. Assign ASVs on forward and reverse reads, merge them, and then once you have the ASV table from each sequencing run, you can merge them together.

But wait: *What contributes to sequencing error in different sequencing runs and why do we need to model errors separately per run with `learnErrors()` in dada2?* Remember the core principles of how Illumina sequencing works! Some things that contribute to this are:  

a. Each sequencing run has a **unique timing for when clusters go out of sync** (*e.g.,* the drop in quality at end of reads, which is typical of Illumina sequencing. This decrease in read quality is especially true of older Illumina sequencing data and is improving with more recent technologies. 
b. The **cluster density is impossible to replicate**. Therefore, the cluster density (and therefore sequence quality) will always be different between sequencing runs (even if it's the same person/samples/sequencing facility!). This is getting easier to replicate with the newer flow cells (*e.g.,* patterned flow cells) but is not yet possible. 
c. **PhiX spike-in will also vary between runs**, even if we try to make it the same! Therefore, the amount of heterogeneity on the flow cell will also be different, which further impacting the sequencing quality.  
d. **Different locations on the flow cell can be impacted differently between runs.** Perhaps an air bubble can get in. Or, as mentioned above, the cluster density is higher and/or lower on a different run/flow cell. In the old, non-patterned flow cells, sometimes clusters can form nearby each other. However, this has been solved on the new, patterned flow cells where clusters have a specific distance between each other. 

**Second, this step uses our quality scores, which is rapidly changing with new sequencing technologies!** Therefore, we need to be mindful of how we run our error models. The traditional error models were developed based on the 40 Phred scores that were generated by early Illumina Sequencing technologies. However, with the new binned quality scores, we need to incorporate new ways of error modeling and this is constanstly changing right now, which you can see in this [very active discussion on the DADA2 GitHub page](https://github.com/benjjneb/dada2/issues/1307). 

Now, let's consider the two types of Illumina data output and how it might inform our error modeling. 

## Learn the Errors

### MiSeq Runs: 40 Phred Scores

Here, in the data we have sequenced here, we have Illumina MiSeq data with the traditional 40 Phred scores. Therefore, we can use the traditional `learnErrors()` command, which we will do below. 

```{r learn-errors, fig.width=12, fig.height=8}
# Forward reads 
error_forward_reads <- 
  learnErrors(filtered_forward_reads, multithread = 20)
# Plot Forward  
forward_error_plot <- 
  plotErrors(error_forward_reads, nominalQ = TRUE) + 
  labs(title = "Forward Read Error Model")

# Reverse reads 
error_reverse_reads <- 
  learnErrors(filtered_reverse_reads, multithread = 20)
# Plot reverse
reverse_error_plot <- 
  plotErrors(error_reverse_reads, nominalQ = TRUE) + 
  labs(title = "Reverse Read Error Model")

# Put the two plots together
forward_error_plot + reverse_error_plot
```

The above plot represents the error rates for each possible transition (A→C, A→G, …) in the forward reads (on the left) and the reverse reads (on the right).

Details of the plots above: 
- **Points**: The observed error rates for each consensus quality score.  
- **Black line**: Estimated error rates after convergence of the machine-learning algorithm.  
- **Red line:** The error rates expected under the nominal definition of the Q-score.  

Similar to what is mentioned in the [DADA2 tutorial](https://benjjneb.github.io/dada2/tutorial_1_8.html): the estimated error rates (black line) are a "reasonably good" fit to the observed rates (points), and the error rates drop with increased quality as expected.  We can now infer ASVs! 


### NextSeq2000 Runs: Binned Quality Scores 

*An important consideration!* There has been active discussion that the **newer binned quality scores from more recent Illumina technologies require many more bases to tune the error models** than the continuous quality scores from MiSeq to prevent spurious ASVs. For example, it's strongly recommended that you play around with and *tune* your number of bases, however, using `1e10` is recommended for the class, as it's likely this will give you enough data to model your error rates. 

Before running your error rates here, one final consideration could be to remove blanks from the error model, which may have unique error rates. 

If you work with binned quality scores, please read through this [DADA2 GitHub Issue](https://github.com/benjjneb/dada2/issues/1307). Specifically, there is a new note from **December 5th, 2024** (*note how recent this is!*) where the developers of DADA2 have proposed a new function to be included. You can also see in this issue that a community member has contributed a loess model to help tune the error model called `loessErrfun_mod`, which my lab has used in the past (since it was before December 5th, 2024)

Now, if you have binned quality scores, the DADA2 authors are suggesting the following for the error models:  

```{r NextSeq-ErrorModel, eval = FALSE, echo = TRUE}
# Inform the model of the specific bins from the Illumina Sequencing run
binnedQs <- c(2, 11, 25, 37)

# Now, use those bins to create the Error Function 
binnedQualErrfun <- makeBinnedQualErrfun(binnedQs)

# Finally, we can actually learn the errors 
learnErrors(filts, errorEstimationFunction=binnedQualErrfun, multi=TRUE)
```

# Infer ASVs

**An important note:** What makes DADA2 so nice is that it infers ASVs separately on the forward and the reverse reads and then merges them afterwards. This is quite a different approach from how OTUs are identified in Mothur and also from UCHIME, oligotyping, and other OTU, MED, and ASV approaches.

```{r infer-ASVs}
# Infer ASVs on the forward sequences
dada_forward <- dada(filtered_forward_reads,
                     err = error_forward_reads, 
                     multithread = TRUE)

typeof(dada_forward)
# Grab a sample and look at it 
dada_forward$`20211005-MA-CWS1P_R1_filtered.fastq.gz`


# Infer ASVs on the reverse sequences 
dada_reverse <- dada(filtered_reverse_reads,
                     err = error_reverse_reads,
                     multithread = TRUE)

# Inspect 
dada_reverse[1]
dada_reverse[30]
```


# Merge Forward & Reverse ASVs

Now, that we have identified our ASVs separately on both the forward and the reverse reads, let's **merge** them together into contiguous (*aka* "contigs") ASVs.

```{r merge-ASVs}
# merge forward and reverse ASVs
merged_ASVs <- mergePairs(dada_forward, filtered_forward_reads, 
                          dada_reverse, filtered_reverse_reads,
                          verbose = TRUE)

# Evaluate the output 
typeof(merged_ASVs)
length(merged_ASVs)
names(merged_ASVs)

# Inspect the merger data.frame from the 20210602-MA-ABB1P 
head(merged_ASVs[[3]])
```

# Create Raw ASV Count Table 
```{r generate-ASV-table, fig.width=3.5, fig.height=3}
# Create the ASV Count Table 
raw_ASV_table <- makeSequenceTable(merged_ASVs)

# Write out the file to data/01_DADA2
write.table(raw_ASV_table, file = "data/01_DADA2/raw_ASV_counts.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)

# Check the type and dimensions of the data
dim(raw_ASV_table)
class(raw_ASV_table)
typeof(raw_ASV_table)
```


# Assess ASV Quality 

## What are the current ASV lengths? 

First, let's inspect the distribution of the ASV sequence lengths across all of the ASVs that we have inferred. First, we need to consider: **What is the length of ASV that expect?** 

Here, we have sequenced the 16S rRNA V4 hypervariable region with the 515F and 806R primers. Therefore: 

1. *What's the total length of our starting amplicons?* 
    - Our primers are named after the location they bind to on the 16S gene. So: 
    - `806 - 515 = 291` 
    - The total length of the starting amplicon is 291 base pairs, with primers.
2. *What is the ASV length without primers?* 
    - The 515F (`GTGYCAGCMGCCGCGGTAA`) primer is 19 base pairs. 
    - The 806R (`GGACTACNVGGGTWTCTAAT`) primer is 20 base pairs. 
    - `291 - 19 - 20 = 252` 
    - The expected ASV length without primers is **252 base pairs**. 
    - **IMPORTANT:** Remember, we did not sequence the primers in our current example, because the [Kozich et al., 2013](https://journals.asm.org/doi/10.1128/aem.01043-13) library prep was used. However, if the regular Illumina 2-step library preparation approach was used, the primers will be sequenced!
3. *What is the length of the trimmed ASVs after `filterAndTrim()`?* 
    - In our `01_QualityTrimming.Rmd`, we used `trimLeft = 8` to remove the bases at the beginning of the forward read and `truncLen = c(250, 248)`, which kept the full forward read length and removed 2 bases at the end of the reverse read. Therefore: 
    - `252 - 8 = 244`
4. *What is the overlap between our forward and reverse reads?*
    - We performed 2x250 paired-end Illumina MiSeq Sequencing. However, from our multiQC report, we can see that our read length is actually 251 bp. 
    - Length of unique forward ASV sequence: `251 - 19 = 232`
    - Length of unique reverse ASV sequence: `251 - 20 = 231`
    - If we have a total read length of 252 base pairs, then our overlap will be `232/251 = 0.924` or 92.4% between our forward and reverse ASVs. 

### ASV Length Stats 

```{r ASV-lengths-stats}
# Calculate summary stats
maxLength_ASV <- max(nchar(getSequences(raw_ASV_table))) # Longest ASV?
minLength_ASV <- min(nchar(getSequences(raw_ASV_table))) # Shortest ASV?
meanLength_ASV <- mean(nchar(getSequences(raw_ASV_table))) # Mean ASV length?
medianLength_ASV <- median(nchar(getSequences(raw_ASV_table))) # Median ASV length?

# Create a table to Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table)))
```

In the table above, we can see that we actually have quite the ASV sequence length distribution! There are some longer ASVs that are clearly spurious! We need to be skeptical! Here's some of the stats that we've calculated!

- Max ASV length: `r maxLength_ASV` base pairs
- Min ASV length: `r minLength_ASV` base pairs
- Mean ASV length: `r meanLength_ASV` base pairs
- Median ASV length: `r medianLength_ASV` base pairs

*Let's take a look at what this looks like graphically!*

### ASV Length Plot 

Now, we will plot these ASV lengths to give us a better idea: 

```{r ASV-lengths-plot}
# Inspect the distribution of sequence lengths of all ASVs in data set 
# AFTER TRIM
data.frame(Seq_Length = nchar(getSequences(raw_ASV_table))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(title = "Raw distribution of ASV length",
       y = "Number of ASVs", x = "ASV Sequence Length (bps)")
```

As we saw in the table, the length of the ASVs is mostly 245 base pairs. 


## Trim ASV lengths 

Now that we are pretty convinced that we have some spurious ASVs, we will trim ASVs that are higher or lower than 245 base pairs in this data set. 

**NOTE:** This is another moment in our workflow where we are sub-setting only the "good parts" of our data to ensure the data quality we have at the end is the best of our ability based on the sequencing run. 

```{r trim-ASVs}
# TRIM THE ASVS
# Let's trim the ASVs to only be the right size, which is indeed 245! 

# We will allow for a few 
raw_ASV_table_trimmed <- raw_ASV_table[,nchar(colnames(raw_ASV_table)) == 245]

# Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table_trimmed)))

# What proportion of total ASV sequences are left in the data? 
sum(raw_ASV_table_trimmed)/sum(raw_ASV_table)

# Inspect the distribution of sequence lengths of all ASVs in dataset 
# AFTER TRIM
data.frame(Seq_Length = nchar(getSequences(raw_ASV_table_trimmed))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(title = "Trimmed distribution of ASV length",
       y = "Number of ASVs", x = "ASV Sequence Length (bps)")

```

**Note the peak at 245 is ABOVE 3000 ASVs!** Now, we have ASV lengths that are only 245 bp and we can move forward onto the next quality control step: removing chimeras. 

# Remove Chimeras

## What is a chimera? 

In 16S rRNA gene sequencing, a **chimera** is an artificial DNA sequence that forms when two or more different DNA templates are mistakenly and erroneously fused together during PCR amplification. **Chimeric sequences do not represent real organisms and must be removed during data processing to avoid false species identifications!** Chimeras are natural artifacts in PCR, especially in high-cycle (>25 cycle) PCR reactions, which are typical in library preparation and sequencing workflows. They arise during: 

1. *Incomplete Extension in Early PCR Cycles.* During early PCR cycles, a DNA fragment may partially amplify before the polymerase detaches. This incomplete product remains in the reaction mixture.
2. *Template Switching in Later PCR Cycles.* In subsequent cycles, the polymerase may bind to and extend a different but similar template, rather than the original one. This results in a hybrid (chimeric) DNA molecule made of two unrelated 16S rRNA gene fragments.
3. *There is increased chimera formation in high-diversity samples.* More diverse microbial communities increase the chance of chimeric formation due to similar overlapping sequences in the sample. Longer PCR extension times and too many PCR cycles (>30) increase chimera formation by allowing more template switching.

## Chimeras must be removed 

- **Chimeras create false ASVs:** Chimeric sequences do not belong to any real microbial species, leading to artificial diversity in sequencing results.
- **Chimeras can mislead ecological interpretations** If not removed, they can inflate richness estimates and misrepresent community composition.

Therefore, chimera removal is an essential step in the analysis of 16S sequencing data to improve the accuracy of downstream analyses, such as taxonomic assignment and diversity assessment. It helps to avoid the inclusion of misleading or spurious sequences that could lead to incorrect biological interpretations. 


```{r rm-chimeras, fig.width=3.5, fig.height=3}
# Remove the chimeras in the raw ASV table
noChimeras_ASV_table <- removeBimeraDenovo(raw_ASV_table_trimmed, 
                                           method="consensus", 
                                           multithread=TRUE, verbose=TRUE)

# Check the dimensions
dim(noChimeras_ASV_table)

# What proportion is left of the sequences? 
percRetained_chimerasTrimmed <- sum(noChimeras_ASV_table)/sum(raw_ASV_table_trimmed)
percRetained_chimerasRaw <-sum(noChimeras_ASV_table)/sum(raw_ASV_table)

# Plot it 
data.frame(Seq_Length_NoChim = nchar(getSequences(noChimeras_ASV_table))) %>%
  ggplot(aes(x = Seq_Length_NoChim )) + 
  geom_histogram()+ 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(title = "Trimmed + Chimera Removal distribution of ASV length",
       y = "Number of ASVs", x = "ASV Sequence Length (bps)")
```

**Note the peak at 245 is now BELOW 3000 ASVs!** So, we have removed 241 ASVs, which were chimeras. This retained `r round(percRetained_chimerasTrimmed *100, digits=2)`% of the trimmed ASV counts and only `r round(percRetained_chimerasRaw *100, digits=2)`% of the raw merged ASV counts. 

# Track the read counts

Here, we will look at the number of reads that were lost in the filtering, denoising, merging, and chimera removal. 

```{r track-reads, fig.width=6, fig.height=4}
# A little function to identify number seqs 
getN <- function(x) sum(getUniques(x))

# Make the table to track the seqs 
track <- cbind(sapply(dada_forward, getN),
               sapply(dada_reverse, getN),
               sapply(merged_ASVs, getN),
               rowSums(noChimeras_ASV_table))

head(track)

# Update column names to be more informative (most are missing at the moment!)
colnames(track) <- c("denoisedF", "denoisedR", "merged", "nochim")
rownames(track) <- sample_names

# Generate a dataframe to track the reads through our DADA2 pipeline
track_counts_df <- 
  track %>%
  # make it a dataframe
  as.data.frame() %>%
  rownames_to_column(var = "names")

# Visualize it in table format 
DT::datatable(track_counts_df)

# Plot it!
track_counts_df %>%
  pivot_longer(denoisedF:nochim, names_to = "read_type", values_to = "num_reads") %>%
  mutate(read_type = fct_relevel(read_type, "denoisedF", "denoisedR", "merged", "nochim")) %>%
  ggplot(aes(x = read_type, y = num_reads, fill = read_type)) + 
  geom_line(aes(group = names), color = "grey") + 
  geom_point(shape = 21, size = 3, alpha = 0.8) + 
  scale_fill_brewer(palette = "Spectral") + 
  labs(x = "Filtering Step", y = "Number of Sequences") + 
  theme_bw()
```


# Assign Taxonomy 

Here, we will use the silva database version 138!
```{r assign-tax}
# Classify the ASVs against a reference set using the RDP Naive Bayesian Classifier described by Wang et al., (2007) in AEM
taxa_train <- 
  assignTaxonomy(noChimeras_ASV_table, 
                 "/workdir/in_class_data/taxonomy/silva_nr99_v138.2_toGenus_trainset.fa.gz", 
                 multithread = 200)

# Add the genus/species information 
taxa_addSpecies <- 
  addSpecies(taxa_train, 
             "/workdir/in_class_data/taxonomy/silva_v138.2_assignSpecies.fa.gz")

# Inspect the taxonomy 
taxa_print <- taxa_addSpecies # Removing sequence rownames for display only
rownames(taxa_print) <- NULL
#View(taxa_print)
```


# Prepare data for export

## 1. ASV Table 

Below, we will prepare the following: 

1. Two ASV Count tables: 
      a. **With ASV seqs:** ASV headers include the *entire* ASV sequence 245 bases.
      b. **with ASV names:** This includes re-written and shortened headers like ASV_1, ASV_2, etc, which will match the names in our fasta file below.  
2. `ASV_fastas`: A fasta file that we can use to build a tree for phylogenetic analyses (e.g. phylogenetic alpha diversity metrics or UNIFRAC dissimilarty).  

### Finalize ASV Count Tables 

First, lets save the ASV table with the ASV sequence name and the ASV fasta file

```{r prepare-ASVcount-Seqtable}
# Give headers more manageable names
# First pull the ASV sequences
asv_seqs <- colnames(noChimeras_ASV_table)
asv_seqs[1:5]

# make headers for our ASV seq fasta file, which will be our asv names
asv_headers <- vector(dim(noChimeras_ASV_table)[2], mode = "character")
asv_headers[1:5]

# loop through vector and fill it in with ASV names 
for (i in 1:dim(noChimeras_ASV_table)[2]) {
  asv_headers[i] <- paste(">ASV", i, sep = "_")
}

# intitution check
asv_headers[1:5]

##### Rename ASVs in table then write out our ASV fasta file! 
#View(noChimeras_ASV_table)
asv_tab <- t(noChimeras_ASV_table)
#View(asv_tab)

## Rename our asvs! 
row.names(asv_tab) <- sub(">", "", asv_headers)
asv_tab[1:5, 1:5]
#View(asv_tab)
```


## 2. Taxonomy Table 
```{r prepare-tax-table}
# Inspect the taxonomy table
#View(taxa_addSpecies)

##### Prepare tax table 
# Add the ASV sequences from the rownames to a column 
new_tax_tab <- 
  taxa_addSpecies%>%
  as.data.frame() %>%
  rownames_to_column(var = "ASVseqs") 
head(new_tax_tab)

# intution check 
stopifnot(new_tax_tab$ASVseqs == colnames(noChimeras_ASV_table))

# Now let's add the ASV names 
rownames(new_tax_tab) <- rownames(asv_tab)
head(new_tax_tab)

### Final prep of tax table. Add new column with ASV names 
asv_tax <- 
  new_tax_tab %>%
  # add rownames from count table for phyloseq handoff
  mutate(ASV = rownames(asv_tab)) %>%
  # Resort the columns with select
  dplyr::select(Kingdom, Phylum, Class, Order, Family, Genus, Species, ASV, ASVseqs)

head(asv_tax)

# Intution check
stopifnot(asv_tax$ASV == rownames(asv_tax), rownames(asv_tax) == rownames(asv_tab))
```

# Write Output Files

Now, we will write the files! We will write the following to the `data/01_DADA2/` folder. We will save both as files that could be submitted as supplements AND as .RData objects for easy loading into the next steps into R.:  

1. `ASV_counts.tsv`: ASV count table that has ASV names that are re-written and shortened headers like ASV_1, ASV_2, etc, which will match the names in our fasta file below. This will also be saved as `data/01_DADA2/ASV_counts.RData`.
2. `ASV_counts_withSeqNames.tsv`: This is generated with the data object in this file known as `noChimeras_ASV_table`. ASV headers include the *entire* ASV sequence ~250bps.  In addition, we will save this as a .RData object as `data/01_DADA2/noChimeras_ASV_table.RData` as we will use this data in `analysis/02_Taxonomic_Assignment.Rmd` to assign the taxonomy from the sequence headers.  
3. `ASVs.fasta`: A fasta file output of the ASV names from `ASV_counts.tsv` and the sequences from the ASVs in `ASV_counts_withSeqNames.tsv`. A fasta file that we can use to build a tree for phylogenetic analyses (e.g. phylogenetic alpha diversity metrics or UNIFRAC dissimilarty).  
4. We will also make a copy of `ASVs.fasta` in `data/02_TaxAss_FreshTrain/` to be used for the taxonomy classification in the next step in the workflow.  
5. Write out the taxonomy table
6. `track_read_counts.RData`: To track how many reads we lost throughout our workflow that could be used and plotted later. We will add this to the metadata in `analysis/02_Taxonomic_Assignment.Rmd`.   


```{r save-files}
# Write BOTH the modified and unmodified ASV tables to a file!
# Write count table with ASV numbered names (e.g. ASV_1, ASV_2, etc)
write.table(asv_tab, 
            file = "data/01_DADA2/ASV_counts.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)

# Write count table with ASV sequence names
write.table(noChimeras_ASV_table, 
            file = "data/01_DADA2/ASV_counts_withSeqNames.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)

# Write out the fasta file for reference later on for what seq matches what ASV
asv_fasta <- c(rbind(asv_headers, asv_seqs))
# Save to a file!
write(asv_fasta, "data/01_DADA2/ASVs.fasta")

# SECOND, let's save the taxonomy tables 
# Write the table 
write.table(asv_tax, 
            file = "data/01_DADA2/ASV_taxonomy.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)

# THIRD, let's save to a RData object 
# Each of these files will be used in the next step of our analysis!
# RData objects are for easy loading :) 
save(noChimeras_ASV_table, file = "data/01_DADA2/noChimeras_ASV_table.RData")
save(asv_tab, file = "data/01_DADA2/ASV_counts.RData")

# And save the track_counts_df a R object, which we will merge with metadata information in the next step of the analysis in nalysis/02_Taxonomic_Assignment. 
save(track_counts_df, file = "data/01_DADA2/track_read_counts.RData")
```

# Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

# Session Information 
```{r session-info}
# Ensure reproducibility 
devtools::session_info()
```
