---
title: "Biodiversity Analysis"
author: "Marian Schmidt"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/04_Biodiversity/",
                      fig.align = "center")
```

# Setting the Environment 

## Set the seed 
```{r set-seed}
set.seed(238428)
```

## Load Libraries 
```{r load-packages}
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               install = FALSE)
```

## Load in Data 
```{r load-data}
load("data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq

# Intiution Check 
min(sample_sums(raw_preprocessed_physeq))

# Make metadata dataframe
metadata_df <-
  raw_preprocessed_physeq %>%
  sample_data() %>%
  data.frame()

head(metadata_df)

# Setting colors for stations 
station_colors <- c(
  "Shipping Channel" = "dodgerblue4",
  "Aransas Bay" = "dodgerblue2",
  "Copano West" = "#D9CC3C",
  "Copano East" = "#A0E0BA",
  "Mesquite Bay" = "#00ADA7")
```


# Goals

1. Calculate the Hill Diversity of the samples. 
2. Evaluate the rarefaction curves. 
3. Evaluate the Diversity values. 
4. Makes notes of specific samples and their seq depth. 

# Diversity Calculations with iNEXT 

```{r calc-div}
# prepare input data 
iNEXT_input_df <- 
  raw_preprocessed_physeq %>%
  otu_table() %>%
  data.frame()
# Quick check
dim(iNEXT_input_df)

# Run iNEXT: Calculate the Hill Numbers 
# Note that: Species in ROWS, Samples in COLUMNS 
# Remember to set the seed! 
#iNEXT_data <- iNEXT(iNEXT_input_df, 
#                    q = c(0,1,2), datatype = "abundance")

# Save the file
#save(iNEXT_data, file = "data/04_Biodiversity/iNEXT_data.RData")
```

# Evaluate the Diversity! 
```{r load-div}
load("data/04_Biodiversity/iNEXT_data.RData")
str(iNEXT_data)
typeof(iNEXT_data)
```

# Plot Diversity 
```{r plot-rarefaction}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
# Check
head(color_df)
# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)


# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Fix the names for merging
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X",  replace = "")) %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(station_colors = station_colors,
            station = names(station_colors)),
            by = "station")
```

# Plot Rarefaction with `ggiNEXT`

```{r ggiNEXT, fig.width=8, fig.height=3.5}
# Plot rarefaction! 
# rarefaction/extrapolation curve, type = 1 

# Order q: 
  # 0 = Richness/ Number of Total taxa
  # 1 = Exponential Shannon / Number of "Common" taxa
  # 2 = Inverse Simpson / Number of "Dominant" taxa 

ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  facet_wrap(~Order.q, scales = "fixed") + 
  scale_color_manual(values = iNEXT_color_df$station_colors, guide = FALSE) + 
  scale_fill_manual(values = iNEXT_color_df$station_colors, guide = FALSE) + 
  scale_shape_manual(values = base::rep(17, nsamples(raw_preprocessed_physeq)),
                     guide = FALSE) +
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  theme(legend.position = "none") 
```

Remembering that an Order q of:  

  - 0 = Richness/ Number of Total taxa
  - 1 = Exponential Shannon / Number of "Common" taxa
  - 2 = Inverse Simpson / Number of "Dominant" taxa 
  
*Also note that iNEXT will extrapolate the number of sequences to double the sequencing depth of the sample!*

Finally, let's draw some conclusions from the plot above (we discussed in class & I'll summarize here):  

1. The best gift of a diversity measure: *All three diversity measures are on the same x-axis: The Effective Number of Species.*  
2. As q increases from 0 to 1 to 2, the abundance of the species is weighted more, so the rarefaction curve platueaus earlier. **This is important because it indicates that richess is the most sensitive to sequencing depth!** However, Shannon and even less sensitive is Simpson!  
3. Most of the sample rarefaction curves appear to plateau but we need to look deeper.  
4. There appears to be a sample with much lower richness than the rest of the dataset. Let's keep our eyes on this sample...  


In your analysis, please provide a clear description! 

# Manually plot Diversity 

## Rarefaction
```{r iNEXT-manual, fig.width=6, fig.height=6}
iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # Fix the samples names 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = "")) %>%
  # join with metadata 
  left_join(., metadata_df, by = "names") %>%
  # Add colors to data frame
  left_join(., data.frame(station_colors = station_colors,
                          station = names(station_colors)),
            by = "station") 
# Inspect 
dim(iNEXT_manual_df)
str(iNEXT_manual_df)

# Plot it - Rarefaction Curve 
iNEXT_manual_df %>%
  # Filter out rows that are calcaulted by rarefaction from iNEXT
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot, be sure to group by the samples if using geom_line!
  ggplot(aes(x = m, y= qD, color = station, group = names)) + 
  geom_line() + 
  # Facet with the station to see the samples better 
  facet_grid(Order.q~station, scales = "free") + 
  scale_color_manual(values = station_colors) + 
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  #guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        legend.title = element_blank()) 
```


While the rarefaction curves from `ggiNEXT` agglomerated the data together, it was hard to see individual samples. Above, we can start to tease apart specific samples within each station that appear higher. 

*Please interpret your results!!*

Some take-aways:  

- Richness:  
    - There does not appear to be much difference between the stations in the total richness.   
    - However, there does appear to be an outlier sample in Aransas Bay.  
    
- Simpson's Diversity:  
    - The **Copano West station** has a bimodal distribution with (1) samples that tend to have more dominant taxa/higher diversity with 40-60 dominant species versus (2) samples that have lower diversity around 25-30 dominant taxa).  
    - **Mesquite Bay** appears to have the lowest Simpson's diversity as the plateau is mostly all below ~22 dominant taxa.  
    - There are two samples in the **Shipping Channel** that have much larger Simpson Diversity values compared to other samples. I wonder who those are! 


# Boxplots of Diversity 

Now, let's zoom in on a bit of a better view of the data by visualizing it in a boxplot...  

```{r div-dfs}
# Prettier station labels 
station_names <- c("Copano\nWest", "Copano\nEast","Mesquite\nBay",
                   "Aransas\nBay", "Shipping\nChannel")

names(station_names) <- c("Copano West", "Copano East","Mesquite Bay",
                          "Aransas Bay", "Shipping Channel")

# Make a dataframe
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  left_join(data.frame(station_names = station_names, 
                       station = names(station_names)), 
                       by = "station")

# Check it 
head(obs_div_df)
```


```{r div-boxplot, fig.height = 3.5, fig.width = 9}
# Boxplots by station 
obs_div_df %>%
  ggplot(aes(x = station_names, y = qD, fill = station, color = station)) + 
  facet_wrap(~Order.q, scales = "free") + 
  geom_jitter(size = 2.5) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = station_colors) + 
  scale_fill_manual(values = station_colors) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        legend.title = element_blank())
```

We can draw that:  

- Richness doesn't appear to be different between the stations, however, we might need to run a statistical test to check!  
- **Mesquite Bay** appears to have the lowest Shannon and Simpson diversity, indicating that it has fewer common/dominant taxa compared to other communities.  
- **Copano West** seems to have the highest Shannon and Simpson Diversity *and* the highest amount of variation. Though, we'd also need to follow this up with a test!  

# Environmental Variable Check 

```{r environmental-pairs, fig.width=8, fig.height=8}
head(metadata_df)
# Pull out environmental variables 
env_df <- 
  metadata_df %>%
  dplyr::select(names, water_tempC:DO_mgL)
# inspect
head(env_df)

# plot the correlations
pairs(dplyr::select(env_df, -names), upper.panel = NULL) 
```

# Diversity vs Salinity 

```{r div-vs-salinity, fig.height=3.5, fig.width=6}
# PSU  = practical salinity unit 
# 1 PSU = 1 g of salt per 1,000 grams of water
# 1 PSU = 1 ppt (part per thousand)
# Freshwater = 0-0.5
# brackish = 0.5 - 30
# Marine = >30

iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  ggplot(aes(x = salinity_psu, y = qD)) + 
  facet_wrap(.~Order.q, scales = "free") + 
  geom_point(aes(color = station)) + 
  stat_smooth(method = "lm", formula = y ~poly(x, 2)) + 
  labs(x = "Salinity (PSU)", y = "# of ASVs") + 
  scale_color_manual(values = station_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_blank()) 
```

Here, there appears to be a few different results: 

- There appears to be a negative & linear relationship between salinity and Richness. This indicates that freshwaters have the highest number of ASVs (genearlly around ~150 ASVs) and as we gain more and more salt in the water, there are fewer species (closer to ~125 ASVs)!  
- Taking the Richness relationship above, we also need to note that there seems to be a high amount of variability within the brackish samples, 
- Shannon & Simpson diversity appear to have a "U" shaped diversity curve!  
- The "U" shape indicates that diversity measures that also weight relative abundance (common and dominant taxa) is highest at the two ends of the spectrum... this indicates that there are fewer dominant species in brackish waters.  



# Session Information
```{r session-info}
devtools::session_info()
```
