---
title: "04A Phylogenetic Tree Construction"
author: "Sophia Aredas"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Goals of this file

The goal is to make a phylogenetic tree!

1. Load in preprocessed phyloseq object
2. Create ASV fasta file from the phyloseq object
3. Align the 16S sequences from fasta file with MAFT
4. Create a tree with FastTree2

Input: raw_preprocessed_physeq
Output: MAFT_aligned_ASVs.fasta; ASVs_unrooted.tree 


# Set seed
```{r set-seed}
set.seed(238428)
```


# Load packages & Phyloseq object
```{r load-environment}
# phytools, ggtree, RColorBrewer
pacman::p_load(phytools, ggtree, RColorBrewer, install = FALSE)

# install phytools
# remotes::install_github("liamrevell/phytools")

# install ggtree
# devtools::install_github("GuangchuangYu/ggtree")

# Load physeq 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq
```

# Write ASV Fasta File
```{r create-proprocessed-fasta}
# Pull out ASV sequences and ASV names
asv_seq_df <- 
  raw_preprocessed_physeq@tax_table %>% 
  data.frame() %>% 
  dplyr::select(ASV, ASVseqs)

#View(asv_seq_df)

# Add the ">" to make fasta header
asv_seq_df$ASV <- paste0(">", asv_seq_df$ASV)

#View(asv_seq_df)

# Create fasta object
asv_seq_fasta <- c(rbind(asv_seq_df$ASV, asv_seq_df$ASVseqs))
head(asv_seq_fasta)

# Write to a file 
write(asv_seq_fasta, 
      file = "data/04_Phylogenetic_Tree/preprocessed_ASVs.fasta")
```

# Align the 16S sequences from fasta file with MAFFT

- `engine.opts = '-l'`: give us original terminal base environment.

```{r run-mafft, engine = 'bash', engine.opts = '-l'}
# Write bash code to run MAFT
# First provide the path to MAFT
export PATH=/programs/mafft/bin:$PATH
# change directories to provide the fasta file we made above 
cd data/04_Phylogenetic_Tree/
pwd

# Set a seed  - using same seed as before for consistency 
RANDOM=238428

# Run MAFT
# To test in the shell directly from Rmd 
# mac: command + option + enter 
# Windows: control + alt + enter 
# For now, use default options, note the version 
# MAFFT automatically knows that it's a nucleotide alignment 
/programs/mafft/bin/mafft --auto preprocessed_ASVs.fasta > MAFFT_aligned_ASVs.fasta

# Change back to the project directory 
cd ../../
pwd
```

# FastTree2 
```{r run-fasttree, engine='bash', engine.opts="-l"}
# Provide export path to fasttree 
export PATH=/programs/FastTree-2.1.11:$PATH

# cd into the alignment file folder 
cd data/04_Phylogenetic_Tree/
pwd

# Run Fasttree to generate phylogenetic tree 
# parameters: 
    # -nt = indicates it's a nucleotide alignment
    # -gtr = generalized time reversible substitution model 
    # -fastest speed up the model, reduce memory usage (recommended for datasets that have >50,000)
    # -log = output a log file 
    # input alignment file 
    # specify the output tree file 
FastTree -nt -gtr -fastest -log FastTree.log MAFFT_aligned_ASVs.fasta > ASVs_unrooted.tree

# Change back to the main directory 
cd ../../
pwd 
echo "The working directory is $PWD"
```