---
title: "Biodiversity across a salinity gradient"
author: "Mar Schmidt"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      # Always relevant to the document directory 
                      # send any figure output to this folder 
                      fig.path = "../figures/05_Biodiversity/",
                      warning = FALSE) 
```


# Goals 

This analysis builds a biodiversity analysis off of the Hill Numbers, which is discussed in a paper entitled "[A conceptual guide to measuring species diversity" by Roswell et al., 2021 in Oikos](https://nsojournals.onlinelibrary.wiley.com/doi/epdf/10.1111/oik.07202). (A great paper!)

1. Calculate the Hill Diversity of the samples using the [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613).
2. Evaluate the interpolated and extrapolated rarefaction curves. 
3. Evaluate the diversity values. 
4. Makes notes of specific samples and their seq depth. 
5. Plot various biodiversity metrics across the salinity gradient to *test the scientific question below!* (Remember, we're doing science!)

## Inputs

1. We will need the `phytree_preprocessed_physeq`, which includes a rooted tree (ideally within the archaea) that we created in `analysis/04B_Phylogenetic_Tree_Inspection.Rmd`. 


## Outputs

*This is an exciting analysis!!! We finally get to answer one of our scientific questions!*

1. Calculated biodiversity measure across the samples. 
2. Figures to include in our scientific paper that visualize the data across the scientific question.
3. Statistical tests conveying the message of our scientific question. 


# Scientific Question

## Field site background

### Mission-Aransas NERR

Remember what we were doing before we got all distracted with the details of ASVs, pre-processing, and phylogenetics? We actually conducted a study to evaluate the role that salinity plays in shaping Estuarine Microbial Communities in the Gulf of Mexico at the long-term SWMP stations at the [Mission-Aransas National Estuarine Research Reserve (NERR)](https://marinescience.utexas.edu/research/mission-aransas-reserve#:~:text=The%20Mission%2DAransas%20National%20Estuarine%20Research%20Reserve%20is%20a%20federal,of%20Texas%20Marine%20Science%20Institute).

**What's a NERR?** The [National Oceanic and Atmospheric Administration (NOAA) maintains 30 estuarine research reserves across the country](https://coast.noaa.gov/nerrs/). There are 5 in the Gulf of Mexico, two (soon to be three!) in the Great Lakes, among others across the United States!  

## Question & Hypotheses

How is microbial biodiversity modified across a salinity gradient?

- *Null Hypothesis:* Microbial biodiversity (*e.g.* richness) does not vary along a salinity gradient.
- *Alternative Hypothesis:* Microbial biodiversity (*e.g.* richness) will increase as salt concentration lowers. There is an inverse relationship between richness and salinity.

*Let's get to calculating Biodiversity!*

# Before starting

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed 
```{r set-seed}
# Any number can be chosen 
set.seed(238428)
```

## Load Packages & Colors
```{r load-packages}
# install ggpubr
#if(!require(devtools)) install.packages("devtools")
#devtools::install_github("kassambara/ggpubr")

# Efficiently load packages 
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq, ggpubr,
               install = FALSE)

# Load in colors for plotting between different files
source("code/colors.R")
```

## Load Data files 
```{r load-data}
# Preprocessed phyloseq object: With the phylogenetic tree!
load("data/04_PhylogeneticTree/phytree_preprocessed_physeqs.RData")

archaeal_rooted_physeq

# Intiution Check 
min(sample_sums(archaeal_rooted_physeq))

# Make metadata dataframe
metadata_df <-
  archaeal_rooted_physeq %>%
  sample_data() %>%
  data.frame()

# Take a look!
head(metadata_df)
```



# Biodiversity Hill Numbers with iNEXT


Next, we will calculate the biodiversity with the [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613) package. iNEXT focuses on three measures of biodiversity via the Hill numbers. Hill numbers have the order q: 

1. **Species richness** when q = 0. 
2. **Shannon diversity** when q = 1, representing the exponential of Shannon entropy.
3. **Simpson diversity** when q = 2, also known as the Inverse Simpson Index.

For each diversity measure, [iNEXT package](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12613) uses the observed abundance of species within a sample to compute diversity estimates and the associated 95% confidence intervals for the following two types of rarefaction and extrapolation (R/E):

1. **Sample-size-based (or size-based) R/E curves**: iNEXT computes diversity estimates for
rarefied and extrapolated samples up to an appropriate size (2x the total number of individuals/sequences in that sample). This type of sampling curve plots the diversity estimates with respect to sample size. 
2.**Coverage‐based R/E curves:** iNEXT computes diversity estimates for rarefied and extrapolated samples based on a standardized level of sample completeness (as measured by sample coverage) up to an appropriate coverage value. This type of sampling curve plots the diversity estimates with respect to sample coverage. **NOTE: Coverage is inapplicable to DADA2 datasets, as they lack singletons** (this is also why we can’t use Amy Willis’s breakaway approach). 

```{r calc-div}
# prepare input data 
iNEXT_input_df <- 
  archaeal_rooted_physeq %>%
  otu_table() %>%
  data.frame()

# Quick check
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]

# Run iNEXT: Calculate the Hill Numbers 
# Note that: Species in ROWS, Samples in COLUMNS 
# Remember to set the seed! 
#iNEXT_data <- iNEXT(iNEXT_input_df, 
#                    q = c(0,1,2), datatype = "abundance")

# Save the file
#save(iNEXT_data, file = "data/05_Biodiversity/iNEXT_data.RData")
```


# Evaluate the Diversity! 

iNEXT takes a long time to calculate, so here we are working with previously calculated iNEXT data.

*An important note:* We’ll be using the size_based output. Our lab has previously been suspicious of the Asymptotic Estimator. Therefore we will use the size_based approach here, with the caveat that we will very rigorously confirm that we’ve approached saturation for all of our samples!

Let’s load the samples and their iNEXT data!

```{r load-div}
load("data/05_Biodiversity/iNEXT_data.RData")
str(iNEXT_data)
typeof(iNEXT_data)
```

# Plot Diversity 
```{r plot-rarefaction}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
# Check
head(color_df)
# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)


# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Fix the names for merging
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X",  replace = "")) %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(station_colors = station_colors,
            station = names(station_colors)),
            by = "station")
```

# Plot Rarefaction with `ggiNEXT`

```{r ggiNEXT, fig.width=8, fig.height=3.5}
# Plot rarefaction! 
# rarefaction/extrapolation curve, type = 1 

# Order q: 
  # 0 = Richness/ Number of Total taxa
  # 1 = Exponential Shannon / Number of "Common" taxa
  # 2 = Inverse Simpson / Number of "Dominant" taxa 

ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  facet_wrap(~Order.q, scales = "fixed") + 
  scale_color_manual(values = iNEXT_color_df$station_colors, guide = FALSE) + 
  scale_fill_manual(values = iNEXT_color_df$station_colors, guide = FALSE) + 
  scale_shape_manual(values = base::rep(17, nsamples(archaeal_rooted_physeq)),
                     guide = FALSE) +
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  theme(legend.position = "none") 
```

Remembering that an Order q of:  

  - 0 = Richness/ Number of Total taxa
  - 1 = Exponential Shannon / Number of "Common" taxa
  - 2 = Inverse Simpson / Number of "Dominant" taxa 
  
*Also note that iNEXT will extrapolate the number of sequences to double the sequencing depth of the sample!*

Finally, let's draw some conclusions from the plot above (we discussed in class & I'll summarize here):  

1. The best gift of a the Hill Numbers: *All three diversity measures are on the same x-axis: The Effective Number of Species or ASVs.*  
2. As q increases from 0 to 1 to 2, the abundance of the species is weighted more, so the rarefaction curve plateaus earlier. **This is important because it indicates that richess is the most sensitive to sequencing depth!** However, Shannon and even less sensitive is Simpson!  
3. Most of the sample rarefaction curves appear to plateau but we need to look deeper.  
4. There appears to be a sample with much lower richness than the rest of the dataset. Let's keep our eyes on this sample...  

Ok, this plot is nice, but since it's autogenerated with the `ggiNEXT` function, I'd like some more freedom to plot. Let's manually plot the data that we pull out from the iNEXT data.

# Manually plot Diversity 

## Rarefaction
```{r iNEXT-manual, fig.width=6, fig.height=6}
iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # Fix the samples names 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = "")) %>%
  # join with metadata 
  left_join(., metadata_df, by = "names") %>%
  # Add colors to data frame
  left_join(., data.frame(station_colors = station_colors,
                          station = names(station_colors)),
            by = "station") %>%
  mutate(station = factor(station, levels = c("Copano West", "Copano East", 
                                              "Mesquite Bay", "Aransas Bay", 
                                              "Shipping Channel")))
# Inspect 
dim(iNEXT_manual_df)
str(iNEXT_manual_df)

# Plot it - Rarefaction Curve 
iNEXT_manual_df %>%
  # Filter out rows that are calcaulted by rarefaction from iNEXT
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot, be sure to group by the samples if using geom_line!
  ggplot(aes(x = m, y= qD, color = station, group = names)) + 
  geom_line() + 
  # Facet with the station to see the samples better 
  facet_grid(Order.q~station, scales = "free") + 
  scale_color_manual(values = station_colors) + 
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  #guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        legend.title = element_blank()) 
```


While the rarefaction curves from `ggiNEXT` agglomerated the data together, it was hard to see individual samples. Above, we can start to tease apart specific samples within each station that appear higher. 

**Some initial take-aways:**  

- *Richness:*  
    - There does not appear to be much difference between the stations in the total richness.   
    - However, there does appear to be an outlier sample in Aransas Bay.  
- *Simpson's Diversity:* 
    - The **Copano West station** has a bimodal distribution with (1) samples that tend to have more dominant taxa/higher diversity with 40-60 dominant species versus (2) samples that have lower diversity around 25-30 dominant taxa).  
    - **Mesquite Bay** appears to have the lowest Simpson's diversity as the plateau is mostly all below ~22 dominant taxa.  
    - There are two samples in the **Shipping Channel** that have much larger Simpson Diversity values compared to other samples. I wonder who those are! 
- *Shannon Diversity:*
    - Here, the trends that were starting to be witnessed in the Shannon diversity are much more amplified!
    
However, all of these changes are qualitative... let's actually quantify by running some stats!

# Boxplots of Diversity 

Now, let's zoom in on a bit of a better view of the data by visualizing it in a boxplot...  

```{r div-dfs}
# Prettier station labels 
station_names <- c("Copano\nWest", "Copano\nEast","Mesquite\nBay",
                   "Aransas\nBay", "Shipping\nChannel")

names(station_names) <- c("Copano West", "Copano East","Mesquite Bay",
                          "Aransas Bay", "Shipping Channel")

# Make a dataframe
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  left_join(data.frame(station_names = station_names, 
                       station = names(station_names)), 
                       by = "station") %>%
  mutate(station_names = factor(station_names, 
                                levels = c("Copano\nWest", "Copano\nEast", 
                                           "Mesquite\nBay", "Aransas\nBay", 
                                           "Shipping\nChannel")))

# Check it 
head(obs_div_df)
```


```{r div-boxplot, fig.height = 3.5, fig.width = 9}
# Boxplots by station 
obs_div_df %>%
  ggplot(aes(x = station_names, y = qD, fill = station, color = station)) + 
  facet_wrap(.~Order.q, scales = "free") + 
  geom_jitter(size = 2.5) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = station_colors) + 
  scale_fill_manual(values = station_colors) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        legend.title = element_blank())
```

We can draw that:  

- Richness doesn't appear to be different between the stations, however, we might need to run a statistical test to check!  
- **Mesquite Bay** appears to have the lowest Shannon and Simpson diversity, indicating that it has fewer common/dominant taxa compared to other communities.  
- **Copano West** seems to have the highest Shannon and Simpson Diversity *and* the highest amount of variation. Though, we'd also need to follow this up with a test!  

## Richness - Statistical Testing

```{r richness-stats}
# Make a richness data frame
richness_df <- 
  obs_div_df %>%
  dplyr::filter(Method == "Observed") %>%
  dplyr::filter(Order.q == 0)

# Add the statistical test! 
kruskal.test(qD ~ station_names, data = richness_df)

# 
my_comparisons <- list( c("Copano\nWest", "Copano\nEast"), c("Copano\nWest", "Mesquite\nBay"), 
                        c("Copano\nWest", "Aransas\nBay"), c("Copano\nWest", "Shipping\nChannel"),
                        c("Copano\nEast", "Mesquite\nBay"), c("Copano\nEast", "Aransas\nBay"),
                        c("Copano\nEast", "Shipping\nChannel"), c("Mesquite\nBay", "Aransas\nBay"),
                        c("Mesquite\nBay", "Shipping\nChannel"), c("Aransas\nBay", "Shipping\nChannel"))

# Now, plot it!
richness_df %>%
  ggplot(aes(x = station_names, y = qD, fill = station, color = station)) + 
  geom_jitter(size = 2.5) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = station_colors) + 
  scale_fill_manual(values = station_colors) + 
  labs(y = "Effective Number of ASVs") + 
  # Add pairwise comparisons p-value
  stat_compare_means(comparisons = my_comparisons,
                     tip.length = 0, hide.ns = TRUE) + 
  stat_compare_means(label.y = 50)  +   # Add global p-value
  theme_bw() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        legend.title = element_blank())
```


# Environmental Variable Check 

```{r environmental-pairs, fig.width=8, fig.height=8}
# Check out metadata
head(metadata_df)

# Pull out environmental variables 
env_df <- 
  metadata_df %>%
  dplyr::select(names, water_tempC:DO_mgL)

# Inspect
head(env_df)

# plot the correlations
pairs(dplyr::select(env_df, -names), upper.panel = NULL) 
```

# Diversity vs Salinity 

## Salinity - what is it? 

PSU  = practical salinity unit 

- 1 PSU = 1 g of salt per 1,000 grams of water
- 1 PSU = 1 ppt (part per thousand)

The gradient is: 

- Freshwater = 0-0.5
- Brackish = 0.5 - 30
- Marine = >30

```{r div-vs-salinity, fig.height=3.5, fig.width=6}
# How does biodiversity change over the salinity gradient
iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  ggplot(aes(x = salinity_psu, y = qD)) + 
  facet_wrap(.~Order.q, scales = "free") + 
  geom_point(aes(color = station)) + 
  stat_smooth(method = "lm", formula = y ~poly(x, 2)) + 
  labs(x = "Salinity (PSU)", y = "# of ASVs") + 
  scale_color_manual(values = station_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_blank()) 
```


Here, there appears to be a few different results: 

- There appears to be a negative & linear relationship between salinity and Richness. This indicates that freshwaters have the highest number of ASVs (genearlly around ~150 ASVs) and as we gain more and more salt in the water, there are fewer species (closer to ~110 ASVs)! We will test this more below. :)  
- Taking the Richness relationship above, we also need to note that there seems to be a high amount of variability within the brackish samples, 
- Shannon & Simpson diversity appear to have a "U" shaped diversity curve!  
- The "U" shape indicates that diversity measures that also weight relative abundance (common and dominant taxa) is highest at the two ends of the spectrum... this indicates that there are fewer dominant species in brackish waters.  

### Explore Richness More

```{r richness-vs-salinity}
# Make a richness data frame
richness_df <- 
  obs_div_df %>%
  dplyr::filter(Method == "Observed") %>%
  dplyr::filter(Order.q == 0)

# Actually run the linear model
rich_vs_sal_lm <- 
  lm(qD ~ salinity_psu, data = richness_df)

# Show me the model results
rich_vs_sal_lm

# Type?
class(rich_vs_sal_lm)

# Summary of lm
summary(rich_vs_sal_lm)
```

In the output:

- **Coefficients** refer to the 𝛽’s
- **Estimate** is the estimate of each coefficient
- **Std. Error** is the standard error of the estimate
- **t value** is the coefficient divided by its standard error
- **Pr(>|t|)** is the p-value for the coefficient
- The **residual standard error** is the estimate of the variance of 𝜖
- **Degrees of freedom** is the sample size minus # of coefficients estimated
- **R-squared** is (roughly) the proportion of variance in the outcome explained by the model, instead we should always report the **adjusted R-squared**, which takes into penalizes the addition of unnecessary predictors, which provides a more accurate measure of model fit.
    - Note that the R-squared value will always increase/stay the same when you add more predictors, even if they are not relevant!
- The **F-statistic** compares the fit of the model as a whole to the null model (with no covariates)
    - When comparing models, we always want to take the model with the higher F-statistic because it indicates a stronger relationship between the predictor varaibles and the dependent variables, suggesting that the model is better. 


```{r plot-richness-sal, fig.height=4, fig.width=4, warning = FALSE}
richness_df %>% 
  ggplot(aes(x = salinity_psu, y = qD)) + 
  geom_point(aes(color = station)) + 
  stat_smooth(method = "lm") + 
  labs(x = "Salinity (PSU)", y = "# of ASVs") + 
  scale_color_manual(values = station_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_blank()) + 
  geom_label(aes(x = 20, y = 70), hjust = 0, 
             label = paste("Adj R2 = ",signif(summary(rich_vs_sal_lm)$adj.r.squared, 2),
                           "\nIntercept =",signif(rich_vs_sal_lm$coef[[1]],3),
                           " \nSlope =",signif(rich_vs_sal_lm$coef[[2]], 2),
                           " \nP =",signif(summary(rich_vs_sal_lm)$coef[2,4], 2)))
```



# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```

